## Setup
```{r setup}
## Confirm libraries
packages = list(
    'readr', 'tidyr', 'dplyr', 'magrittr',
    'HDF5Array', 'SummarizedExperiment', 'S4Vectors', 'DESeq2',
    'creater', 'ggplot2', 'ggridges', 'ggupset', 'patchwork',
    'foreach', 'glmnet', 'caret', 'pROC',
    'topGO', 'org.Hs.eg.db'
)
has_packages <- unlist(lapply(packages, function(p) {p %in% rownames(installed.packages())}))
if (!all(has_packages)) {
    stop(paste0(
        "Missing required packages:\n",
        paste(packages[!has_packages], collapse = "\n")
    ))
}

## Operators
`%>%` <- tidyr::`%>%`
`%do%` <- foreach::`%do%`
`%:%` <- foreach::`%:%`

## Set default printing format
knitr::opts_chunk$set(
    fig.align = 'center',
    fig.path = 'notebooks/figures/',
    fig.width = 8,
    cache = TRUE,
    messages = FALSE,
    warning = FALSE
)

## Color scheming
barretts_palette <- c(
    "#666666", "#D95F02",
    "#1B9E77", "#E6AB02", "#E7298A",
    "#7570B3", "#A6761D", "#2C9DEF"
) %>%
    magrittr::set_names(c(
        "Protein coding", "lncRNA",
        "Pseudogene", "rRNA", "tRNA",
        "Mitochondrial", "Other", "Novel Gene"
    ))

theme_cell <- function() {
    return(
        ggthemes::theme_foundation(
            base_size = 6,
            base_family = 'Arial'
        ) +
        ggplot2::theme(
            plot.title = ggplot2::element_text(
                size = 6,
                color = "#282A36",
                face = 'bold',
                hjust = 0,
                vjust = 0.5
            ),
            plot.background = ggplot2::element_blank(),
            plot.tag = ggplot2::element_text(
                size = 10,
                face = "bold",
                color = "#282A36"
            ),
            plot.margin = ggplot2::margin(1, 1, 1, 1, unit = "pt"),
            panel.background = ggplot2::element_blank(),
            panel.border = ggplot2::element_blank(),
            panel.grid.major = ggplot2::element_blank(),
            panel.grid.minor = ggplot2::element_blank(),
            axis.line = ggplot2::element_line(color = "#282A36"),
            axis.line.x = NULL,
            axis.line.y = NULL,
            axis.text = ggplot2::element_text(size = 6, color = "#282A36"),
            axis.text.x = ggplot2::element_text(
                margin = ggplot2::margin(t = 0.5),
                vjust = 1, color = "#282A36"
            ),
            axis.text.x.top = ggplot2::element_text(
                margin = ggplot2::margin(b = 0.5),
                vjust = 0, color = "#282A36"
            ),
            axis.text.y = ggplot2::element_text(
                margin = ggplot2::margin(r = 0.25),
                hjust = 1, color = "#282A36"
            ),
            axis.text.y.right = ggplot2::element_text(
                margin = ggplot2::margin(l = 0.25),
                hjust = 0, color = "#282A36"
            ),
            axis.ticks = ggplot2::element_line(),
            axis.ticks.length = ggplot2::unit(2, "pt"),
            axis.ticks.length.x = NULL,
            axis.ticks.length.x.top = NULL,
            axis.ticks.length.x.bottom = NULL,
            axis.ticks.length.y = NULL,
            axis.ticks.length.y.left = NULL,
            axis.ticks.length.y.right = NULL,
            strip.text = ggplot2::element_text(
                size = 6, face = "bold", color = "#282A36"
            ),
            strip.background = ggplot2::element_blank(),
            strip.placement = "outside",
            legend.spacing = ggplot2::unit(0, "pt"),
            legend.title = ggplot2::element_text(
                size = 6, face="italic", color = "#282A36"
            ),
            legend.text = ggplot2::element_text(
                size = 6, face="bold", color = "#282A36"
            ),
            legend.justification = c(0.1, 0.75),
            legend.box.just = "right",
            legend.box.spacing = ggplot2::unit(-0.25, "pt"),
            legend.position = "right",
            legend.background = ggplot2::element_blank()
        )
    )
}

# Analysis parameters
pval_thresh <- 0.05
logFC_thresh <- 1
min_lib_size <- 1e5
min_gene_detection <- 1e4
min_expr_cutoff <- 3
min_expr_frac <- 0.25
```

## Data
```{r data}
## Read in H5 files
novel_se <- creater::load_counts("data/counts/short_read/novel")
gencode_se <- creater::load_counts("data/counts/short_read/gencode")

## Read in metadata
metadata <- "data/metadata.csv" %>%
    readr::read_csv(show_col_types = FALSE, progress = FALSE) %>%
    dplyr::mutate(
        condition = dplyr::case_when(
            condition == "healthy" ~ "Control",
            condition == "barretts" ~ "Barrett's Esophagus",
            condition == "adenocarcinoma" ~ "Esophageal Adenocarcinoma"
        ),
        condition = factor(condition, levels = c(
            "Control",
            "Barrett's Esophagus",
            "Esophageal Adenocarcinoma"
        )),
        condition_code = dplyr::case_when(
            condition == "Control" ~ "CT",
            condition == "Barrett's Esophagus" ~ "BE",
            condition == "Esophageal Adenocarcinoma" ~ "EA"
        ),
        sample_id = make.unique(condition_code, sep=""),
        condition_code = factor(condition_code, levels = c("CT", "BE", "EA")),
        sex = factor(sex, levels = c('F', 'M')),
        age = as.numeric(age)
    ) %>%
    dplyr::mutate(
        age = dplyr::case_when(
            is.na(age) ~ mean(age, na.rm = TRUE),
            TRUE ~ age
        ),
        .by = c("condition", "sex")
    ) %>%
    dplyr::mutate(
        scaled_age = (age - mean(age)) / sd(age),
    )
SummarizedExperiment::colData(novel_se) <- dplyr::left_join(
        data.frame(
            names = colnames(novel_se),
            sample = stringr::str_remove(colnames(novel_se), "_novel$")
        ),
        metadata,
        by = "sample"
    ) %>%
    S4Vectors::DataFrame()
colnames(novel_se) <- SummarizedExperiment::colData(novel_se)$names
SummarizedExperiment::colData(gencode_se) <- dplyr::left_join(
        data.frame(
            names = colnames(gencode_se),
            sample = stringr::str_remove(colnames(gencode_se), "_gencode$")
        ),
        metadata,
        by = "sample"
    ) %>%
    S4Vectors::DataFrame()
colnames(gencode_se) <- SummarizedExperiment::colData(gencode_se)$names

## Read in biotype annotations
biotypes <- readr::read_csv(
            "/private/groups/kimlab/alex/proj/barretts/LSR-DISC/data/metadata/gene_info.csv",
            col_names = c("header", "gene_name"),
            show_col_types = FALSE, progress = FALSE
        ) %>%
        tidyr::separate_wider_delim(
            cols = 'header',
            delim = "|",
            names = c(
                "tx_id", "gene_id", "gid2",
                "txid2", "tx_name", "gene_name2",
                "length", "tx_biotype", "blank"
            )
        ) %>%
        dplyr::select(gene_id, gene_name, tx_biotype) %>%
        dplyr::summarize(
            .by = c("gene_id", "gene_name"),
            gene_biotype = dplyr::case_when(
                any(startsWith(gene_name, "MT-")) ~ "Mitochondrial",
                any(tx_biotype == "protein_coding") ~ "Protein coding",
                any(tx_biotype == "lncRNA") ~ "lncRNA",
                any(unlist(lapply(tx_biotype, function(x) {
                    grepl("tRNA", x)
                }))) ~ "tRNA",
                any(unlist(lapply(tx_biotype, function(x) {
                    grepl("rRNA", x)
                }))) ~ "rRNA",
                any(unlist(lapply(tx_biotype, function(x) {
                    grepl("pseudogene", x)
                }))) ~ "Pseudogene",
                TRUE ~ "Other"
            )
        ) %>%
        dplyr::bind_rows(
            data.frame(
                gene_id = rownames(novel_se) %>%
                    magrittr::extract(startsWith(., "novel_gene_")),
                gene_name = rownames(novel_se) %>%
                    magrittr::extract(startsWith(., "novel_gene_")) %>%
                    stringr::str_replace("novel_gene_", "Novel gene ") %>%
                    stringr::str_replace("_", "."),
                gene_biotype = "Novel Gene"
            )
        )
SummarizedExperiment::rowData(novel_se) <- dplyr::left_join(
        as.data.frame(SummarizedExperiment::rowData(novel_se)) %>%
            dplyr::mutate(gene_id = rownames(.)),
        biotypes,
        by = "gene_id"
    ) %>%
    S4Vectors::DataFrame()
SummarizedExperiment::rowData(gencode_se) <- dplyr::left_join(
        as.data.frame(SummarizedExperiment::rowData(gencode_se)) %>%
            dplyr::mutate(gene_id = rownames(.)),
        biotypes,
        by = "gene_id"
    ) %>%
    S4Vectors::DataFrame()

## Merge novel and gencode quants
gene_list <- rownames(novel_se)[
    (rownames(novel_se) %in% rownames(gencode_se)) |
    startsWith(rownames(novel_se), "novel_gene_")
]
combined_se <- SummarizedExperiment::SummarizedExperiment(
    assays = lapply(SummarizedExperiment::assayNames(novel_se), function (name) {
            novel <- SummarizedExperiment::assay(novel_se[gene_list,], name) %>%
                as.data.frame() %>%
                magrittr::set_colnames(paste0(colnames(.), "_novel")) %>%
                dplyr::mutate(gene_id = rownames(.))
            gencode <- SummarizedExperiment::assay(gencode_se[rownames(gencode_se) %in% gene_list,], name) %>%
                as.data.frame() %>%
                magrittr::set_colnames(paste0(colnames(.), "_gencode")) %>%
                dplyr::mutate(gene_id = rownames(.))
            combined <- dplyr::left_join(
                novel,
                gencode,
                by = "gene_id",
                suffix = c("_novel", "_gencode")
            ) %>%
            dplyr::mutate(dplyr::across(
                dplyr::starts_with("SLDK"),
                ~ tidyr::replace_na(., 0)
            )) %>%
            dplyr::mutate(dplyr::across(
                dplyr::starts_with("SLDK"),
                ~ floor(as.numeric(.))
            )) %>%
            dplyr::mutate(
                gene_id = factor(gene_id, levels = names(SummarizedExperiment::rowRanges(novel_se[gene_list,])))
            ) %>%
            dplyr::arrange(gene_id) %>%
            magrittr::set_rownames(.$gene_id) %>%
            dplyr::select(-gene_id)
        }) %>%
        magrittr::set_names(SummarizedExperiment::assayNames(novel_se)),
    colData = data.frame(names = c(paste0(metadata$sample, "_novel"), paste0(metadata$sample, "_gencode"))) %>%
        dplyr::mutate(
            sample = stringr::str_remove(names, "_novel$|_gencode$"),
            transcriptome = dplyr::case_when(
                endsWith(names, "_novel") ~ "Novel cfRNA + GENCODE",
                endsWith(names, "_gencode") ~ "GENCODE"
            )
        ) %>%
        dplyr::left_join(
            metadata,
            by = "sample"
        ) %>%
        S4Vectors::DataFrame(),
    rowRanges = SummarizedExperiment::rowRanges(novel_se[gene_list,])
)
SummarizedExperiment::rowData(combined_se) <- SummarizedExperiment::rowData(novel_se[gene_list,])

## Filter bad samples and low expression genes
rna_filter <- function(se, min_lib_size = 1e5, min_gene_detection = 1e4, min_expr_cutoff = 3, min_expr_frac = 0.25) {
    failed_samples <- colSums(SummarizedExperiment::assay(se)) <
        min_lib_size
    low_coverage_samples <- (SummarizedExperiment::assay(se) >= 1) %>%
        as.data.frame() %>%
        dplyr::filter(!startsWith(rownames(.), "novel_gene_")) %>%
        colSums() %>%
        { . < min_gene_detection }
    bad_samples <- failed_samples | low_coverage_samples

    bad_genes <- SummarizedExperiment::assay(se[,!bad_samples]) %>%
        { . > min_expr_cutoff } %>%
        rowSums() %>%
        { . < (min_expr_frac * ncol(se[,!bad_samples])) }

    return(list(bad_genes, bad_samples))
}

novel_filter <- rna_filter(novel_se)
cat(paste0(
    "Filter removed ", sum(novel_filter[[2]]), " samples and ",
    sum(novel_filter[[1]]), " genes from novel_se.\n")
)

if (!file.exists('results/models/novel_dds.rds')) {
    cat("Running DESeq2...\n")
    novel_dds <- DESeq2::DESeqDataSet(
            novel_se[!novel_filter[[1]], !novel_filter[[2]]],
            design = ~condition_code + sex + scaled_age
        ) %>%
        DESeq2::DESeq(quiet = TRUE)
    readr::write_rds(novel_dds, "results/models/novel_dds.rds")
} else {
    cat("Loading previous DESeq2 run...\n")
    novel_dds <- readr::read_rds("results/models/novel_dds.rds")
}
```


# Figure 1 - Data quality

# Panel E-G: Gene detection by transcriptome and biotype
```{r fig_1.panel_bcd}
fig1.panelBCD <- lapply(list('Protein coding', 'lncRNA', 'Novel Gene'), function(biotype) {
    plot <- combined_se %>%
        SummarizedExperiment::assay() %>%
        { . > 1 } %>%
        as.data.frame() %>%
        dplyr::mutate(gene_id = rownames(.)) %>%
        dplyr::left_join(biotypes, by = "gene_id") %>%
        dplyr::filter(gene_biotype == biotype) %>%
        tidyr::pivot_longer(dplyr::starts_with("SLDK"), names_to = "names", values_to = "detected") %>%
        dplyr::left_join(
            SummarizedExperiment::colData(combined_se) %>% as.data.frame(),
            by = c("names")
        ) %>%
        dplyr::summarize(ndetected = sum(detected), .by = c('sample', 'condition_code', 'transcriptome')) %>%
        dplyr::mutate(
            transcriptome = factor(
                transcriptome,
                levels = c("GENCODE", "Novel cfRNA + GENCODE")
            ),
            condition_code = factor(
                condition_code,
                levels = c("CT", "BE", "EA")
            )
        ) %>%
        ggplot2::ggplot(ggplot2::aes(x = condition_code, y = ndetected, fill = transcriptome)) +
        ggplot2::geom_boxplot() +
        ggplot2::scale_fill_manual(values = c(
                "GENCODE" = unname(barretts_palette["Protein coding"]),
                "Novel cfRNA + GENCODE" = unname(barretts_palette["Novel Gene"])
        )) +
        ggplot2::labs(
            x = "Sample Condition",
            y = "Number of detected genes",
            fill = "Transcriptome",
            title = biotype
        ) +
        theme_cell() +
        ggplot2::theme(
            legend.position = "bottom",
            plot.title = ggplot2::element_text(hjust = 0.5)
        )
    panels = c(
        'Protein coding' = 'B',
        'lncRNA' = 'C',
        'Novel Gene' = 'D'
    )
    ggplot2::ggsave(
        filename = paste0("results/panels/figure1_panel", panels[[biotype]], ".pdf"),
        plot = plot, width = 5.8, height = 5.8, units = "cm",
        device = cairo_pdf
    )
    ggplot2::ggsave(
        filename = paste0("results/panels/figure1_panel", panels[[biotype]], ".png"),
        plot = plot, width = 5.8, height = 5.8, units = "cm",
        dpi = 600
    )
    return(list(plot))
}) %>%
do.call(c, .) %>%
c(list(patchwork::guide_area())) %>%
patchwork::wrap_plots() +
patchwork::plot_layout(
    guides = 'collect',
    design = 'ABC\nDDD',
    heights = c(3, 0.1)
) +
patchwork::plot_annotation(tag_levels = list(c('B', 'C', 'D')))
```

## Panel E: Novel gene transcript length
```{r fig1.panel_e}
fig1.panelE <- SummarizedExperiment::rowRanges(combined_se) %>%
    as.data.frame() %>%
    dplyr::mutate(
        is_novel = startsWith(group_name, "novel_gene_")
    ) %>%
    dplyr::select(group_name, width, is_novel) %>%
    dplyr::filter( width <= 3000) %>%
    ggplot2::ggplot(ggplot2::aes(x=width, fill=is_novel, y = is_novel)) +
    ggridges::geom_density_ridges(
        alpha = 1, color = 'white',
        scale = 1, rel_min_height = 0.0001,
    ) +
    theme_cell() +
    ggplot2::scale_fill_manual(values = c(
        "TRUE" = unname(barretts_palette["Novel Gene"]),
        "FALSE" = "#666666"
    )) +
    ggplot2::labs(
        x = "Transcript length (bp)",
        y = "Density"
    ) +
    ggplot2::theme(
        legend.position = "none",
        axis.text.y = ggplot2::element_text(angle = 90, vjust = 0.1, hjust = -0.1),
        axis.ticks.y = ggplot2::element_blank()
    ) +
    ggplot2::scale_y_discrete(
        labels = c(
            "FALSE" = "GENCODE genes",
            "TRUE" = "Novel genes"
        ),
        expand = c(0,0)
    ) +
    ggplot2::scale_x_continuous(expand = c(0,0))
ggplot2::ggsave(
    filename = "results/panels/figure1_panelE.pdf",
    plot = fig1.panelE,
    width = 5.8, height = 5.8, units = "cm",
    device = cairo_pdf
)
ggplot2::ggsave(
    filename = "results/panels/figure1_panelE.png",
    plot = fig1.panelE,
    width = 5.8, height = 5.8, units = "cm",
    dpi = 600
)
```

## Panel FGHI: Novel genes per chromosome
```{r fig1.panel_fghi}
novel_enrichment <- DESeq2::results(novel_dds, contrast = c("condition_code", "EA", "CT")) %>%
    as.data.frame() %>%
    dplyr::filter(!is.na(padj)) %>%
    dplyr::mutate(
        gene_id = rownames(.),
        signif = (padj < pval_thresh) & (abs(log2FoldChange) >= logFC_thresh)
    ) %>%
    dplyr::filter(signif) %>%
    dplyr::mutate(enriched = dplyr::case_when(
            log2FoldChange > 0 ~ "EA-enriched",
            log2FoldChange < 0 ~ "CT-enriched"
        )) %>%
    dplyr::left_join(
        biotypes,
        by = "gene_id"
    ) %>%
    dplyr::filter(gene_biotype == "Novel Gene") %>%
    dplyr::bind_rows(
        DESeq2::results(novel_dds, contrast = c("condition_code", "BE", "CT")) %>%
        as.data.frame() %>%
        dplyr::filter(!is.na(padj)) %>%
        dplyr::mutate(
            gene_id = rownames(.),
            signif = (padj < pval_thresh) & (abs(log2FoldChange) >= logFC_thresh),
        ) %>%
        dplyr::filter(signif) %>%
        dplyr::mutate(enriched = dplyr::case_when(
            log2FoldChange > 0 ~ "BE-enriched",
            log2FoldChange < 0 ~ "CT-enriched"
        )) %>%
        dplyr::left_join(
            biotypes,
            by = "gene_id"
        ) %>%
        dplyr::filter(gene_biotype == "Novel Gene")
    ) %>%
    dplyr::summarize(
        .by = gene_id,
        enriched = paste(sort(unique(enriched)), collapse = " & ") %>%
            stringr::str_replace_all('-enriched', '') %>%
            paste0('-enriched')
    )

fig1.panelFGHI <- purrr::imap(list(
        'F' = "BE & EA-enriched",
        'G' = "EA-enriched",
        'H' = "BE-enriched",
        'I' = "CT-enriched"
    ), function(group, panel) {
    plot <- data.frame(gene_id = rownames(combined_se)) %>%
        dplyr::filter(startsWith(gene_id, 'novel_gene_')) %>%
        dplyr::mutate(
            chromosome = stringr::str_replace(gene_id, "novel_gene_", "") %>%
            stringr::str_replace("_.*", "") %>%
            stringr::str_replace("chr", "")
        ) %>%
        dplyr::filter(chromosome != "Un") %>%
        dplyr::left_join(
            novel_enrichment %>%
            dplyr::select(gene_id, enriched),
            by = 'gene_id'
        ) %>%
        dplyr::filter(!is.na(enriched)) %>%
        dplyr::summarize(ngenes = dplyr::n(), .by = c('chromosome', 'enriched')) %>%
        dplyr::bind_rows(
            expand.grid(
                chromosome = c(1:22, "X", "Y"),
                enriched = c("BE & EA-enriched", "EA-enriched", "BE-enriched", "CT-enriched"),
                ngenes = 0
            )
        ) %>%
        dplyr::left_join(data.frame(
            chromosome = c(1:22, "X", "Y"),
            bases = c(
                248956422, 242193529, 198295559, 190214555,
                181538259, 170805979, 159345973, 145138636,
                138394717, 133797422, 135086622, 133275309,
                114364328, 107043718, 101991189, 90338345,
                83257441, 80373285, 58617616, 64444167,
                46709983, 50818468, 156040895, 57227415
            )
        ), by = "chromosome") %>%
        dplyr::mutate(
            ngenes_per_gb = as.numeric(ngenes) / (bases / 1e9),
            chromosome = factor(chromosome, levels = c(as.character(1:22), "X", "Y")),
            enriched = factor(enriched, levels = c(
                "BE & EA-enriched", "EA-enriched", "BE-enriched", "CT-enriched"
            ))
        ) %>%
        dplyr::summarize(
            ngenes_per_gb = sum(ngenes_per_gb),
            .by = c('chromosome', 'enriched')
        ) %>%
        dplyr::filter(enriched == group) %>%
        ggplot2::ggplot(ggplot2::aes(x = chromosome, y = ngenes_per_gb, fill = enriched)) +
        ggplot2::geom_bar(stat = "identity", position = 'dodge') +
        ggplot2::labs(
            x = "Chromosome",
            y = "DE Novel genes / GB",
            fill = "Enrichment",
            title = paste0(group, " novel cfRNAs")
        ) +
        ggplot2::scale_fill_manual(values = list(
            'BE-enriched' = "#2C9DEF",
            'EA-enriched' = "#D95F02",
            'BE & EA-enriched' = "#A6761D",
            'CT-enriched' = "#666666",
            'Not significant' = "lightgray"
        )) +
        theme_cell() +
        ggplot2::scale_y_continuous(expand = c(0,0)) +
        ggplot2::scale_x_discrete(
            breaks = c(
                "1", "", "3", "", "5", "", "7", "", "9", "",
                "11", "", "13", "", "15", "", "17", "19", "",
                "21", "", "X", "Y"
            )
        ) +
        ggplot2::theme(
            legend.position = 'none',
            axis.ticks.x = ggplot2::element_blank()
        )
    ggplot2::ggsave(
        filename = paste0("results/panels/figure1_panel", panel, ".pdf"),
        plot = plot,
        width = 5, height = 3,
        device = cairo_pdf
    )
    ggplot2::ggsave(
        filename = paste0("results/panels/figure1_panel", panel, ".png"),
        plot = plot,
        width = 9.67, height = 5.8, dpi = 600
    )
    return(list(plot))
}) %>%
do.call(c, .) %>%
patchwork::wrap_plots(ncol = 2)
```

## Figure 1
```{r figure1}
fig1.panelA <- ggplot2::ggplot() +
    {
        magick::image_read_pdf('results/panels/figure1_panelA.pdf', pages = 2) %>%
        grid::rasterGrob(
            width = ggplot2::unit(1, "npc"),
            height = ggplot2::unit(1, "npc")
        ) %>%
        ggplot2::annotation_custom()
    } +
    theme_cell()
figure1 <- patchwork::wrap_plots(list(
        patchwork::free(fig1.panelA),
        fig1.panelBCD,
        fig1.panelE,
        fig1.panelFGHI
    )) +
    patchwork::plot_layout(
        guides = 'keep', tag_level = 'keep',
        heights = c(2.8, 3, 3),
        widths = c(1, 3),
        design = 'AA\nBB\nCD',
    ) +
    patchwork::plot_annotation(tag_levels = 'A')
ggplot2::ggsave(
    filename = "results/figures/figure1.pdf",
    plot = figure1,
    width = 17.4, height = 17, units = "cm",
    device = cairo_pdf
)
ggplot2::ggsave(
    filename = "results/figures/figure1.png",
    plot = figure1,
    width = 17.4, height = 17, units = "cm",
    dpi = 600
)
```



# Figure 2 - Differential expression

## Panel A-B: Differential expression results
```{r fig2.panel_ab}
fig2.panelAB <- lapply(list("BE", "EA"), function(numer) {
    de_res <- DESeq2::results(novel_dds, contrast = c("condition_code", numer, "CT")) %>%
        as.data.frame() %>%
        dplyr::filter(!is.na(padj)) %>%
        dplyr::mutate(
            gene_id = rownames(.),
            signif = dplyr::case_when(
                (padj < pval_thresh) & (abs(log2FoldChange) >= logFC_thresh) ~
                    dplyr::case_when(
                        log2FoldChange > 0 ~ paste0(numer, "-enriched"),
                        log2FoldChange < 0 ~ "CT-enriched"
                    ),
                TRUE ~ "Not significant"
            )
        ) %>%
        dplyr::left_join(
            biotypes,
            by = "gene_id"
        ) %>%
        dplyr::mutate(biotype = dplyr::case_when(
            signif == "Not significant" ~ "Not significant",
            TRUE ~ gene_biotype
        )) %>%
        dplyr::bind_rows(data.frame(biotype = unique(.$gene_biotype))) %>%
        dplyr::slice_sample(
            by = signif, n = 1000, replace = FALSE
        ) %>%
        ggplot2::ggplot(ggplot2::aes(x = log2FoldChange, y = -log10(padj), color = biotype)) +
        ggplot2::geom_vline(xintercept = c(-logFC_thresh, logFC_thresh), linetype = "dashed", color = "lightgray") +
        ggplot2::geom_hline(yintercept = -log10(pval_thresh), linetype = "dashed", color = "lightgray") +
        ggplot2::geom_text(
            data = data.frame(
                label = c('CT-enriched', paste0(numer, '-enriched')),
                x = c(-Inf, Inf), y = c(Inf, Inf),
                hjust = c(-0.1, 1.1), vjust = c(-0.1, -0.1)
            ), ggplot2::aes(x = x, y = y, label = label, hjust = hjust, vjust = vjust),
            size = 6, size.unit = "pt", color = "lightgray"
        ) +
        ggplot2::geom_point() +
        ggplot2::scale_color_manual(
            values = c(
                "Not significant" = "lightgray",
                barretts_palette
            ),
            breaks = names(barretts_palette)
        ) +
        ggplot2::labs(x="Log2FoldChange", y="-Log10 padj", color = "Gene biotype") +
        theme_cell() +
        ggplot2::guides(color = ggplot2::guide_legend(
            override.aes = list(
                size = 8, shape=15
            )
        )) +
        ggplot2::theme(
            legend.key.spacing.y = ggplot2::unit(-.16, "cm")
        ) +
        ggplot2::coord_cartesian(clip = "off")
    panel_letter = c("BE" = "A", "EA" = "B")
    ggplot2::ggsave(
        filename = paste0("results/panels/figure2_panel", panel_letter[[numer]], ".pdf"),
        plot = de_res,
        width = 8.7, height = 5, units = "cm",
        device = cairo_pdf
    )
    ggplot2::ggsave(
        filename = paste0("results/panels/figure2_panel", panel_letter[[numer]], ".png"),
        plot = de_res,
        width = 4, height = 4, dpi = 600
    )
    return(list(de_res))
})
```

## Panel C: Disease upset
```{r fig2.panel_c}
fig2.panelC <- dplyr::left_join(
        DESeq2::results(novel_dds, contrast = c("condition_code", "BE", "CT")) %>%
            as.data.frame() %>%
            dplyr::filter(!is.na(padj)) %>%
            dplyr::mutate(
                gene_id = rownames(.),
                BE_signif = (padj < pval_thresh) & (log2FoldChange >= logFC_thresh)
            ) %>%
            dplyr::select(gene_id, BE_signif),
        DESeq2::results(novel_dds, contrast = c("condition_code", "EA", "CT")) %>%
            as.data.frame() %>%
            dplyr::filter(!is.na(padj)) %>%
            dplyr::mutate(
                gene_id = rownames(.),
                EA_signif = (padj < pval_thresh) & (log2FoldChange >= logFC_thresh)
            ) %>%
            dplyr::select(gene_id, EA_signif),
        by = "gene_id"
    ) %>%
    dplyr::filter(BE_signif | EA_signif) %>%
    dplyr::mutate(
        present = dplyr::case_when(
            BE_signif & EA_signif ~ "BE+EA",
            BE_signif ~ "BE",
            EA_signif ~ "EA"
        ),
        present = factor(present, levels = c("BE", "EA", "BE+EA"))
    ) %>%
    dplyr::left_join(
        biotypes %>% dplyr::select(gene_id, gene_biotype),
        by = "gene_id"
    ) %>%
    dplyr::mutate(
        gene_biotype = factor(gene_biotype, levels = names(barretts_palette))
    ) %>%
    dplyr::bind_rows(expand.grid(
        gene_biotype = unique(.$gene_biotype),
        present = factor(c("BE", "EA", "BE+EA"), levels = c("BE", "EA", "BE+EA"))
    )) %>%
    dplyr::summarize(
        .by = c("gene_biotype", "present"),
        count = dplyr::n() - 1
    ) %>%
    ggplot2::ggplot(ggplot2::aes(x = present, y = count, fill = gene_biotype)) +
    ggplot2::geom_bar(stat = "identity", position = 'dodge') +
    ggupset::axis_combmatrix(sep = "\\+") +
    ggplot2::scale_y_continuous(expand = c(0,0)) +
    ggplot2::scale_fill_manual(values = barretts_palette) +
    ggplot2::labs(
        x = "Differentially expressed genes",
        y = "Number of genes",
        fill = "Gene biotype",
        title = "DE genes enriched in disease conditions"
    ) +
    theme_cell() +
    ggplot2::theme(legend.position = "none")
ggplot2::ggsave(
    filename = "results/panels/figure2_panelC.pdf",
    plot = fig2.panelC,
    width = 8.7, height = 5,
    device = cairo_pdf
)
ggplot2::ggsave(
    filename = "results/panels/figure2_panelC.png",
    plot = fig2.panelC,
    width = 8.7, height = 5, dpi = 600
)
```

## Panel D: Control upset
```{r fig2.panel_d}
fig2.panelD <- dplyr::left_join(
        DESeq2::results(novel_dds, contrast = c("condition_code", "BE", "CT")) %>%
            as.data.frame() %>%
            dplyr::filter(!is.na(padj)) %>%
            dplyr::mutate(
                gene_id = rownames(.),
                BE_signif = (padj < pval_thresh) & (log2FoldChange <= -logFC_thresh)
            ) %>%
            dplyr::select(gene_id, BE_signif),
        DESeq2::results(novel_dds, contrast = c("condition_code", "EA", "CT")) %>%
            as.data.frame() %>%
            dplyr::filter(!is.na(padj)) %>%
            dplyr::mutate(
                gene_id = rownames(.),
                EA_signif = (padj < pval_thresh) & (log2FoldChange <= -logFC_thresh)
            ) %>%
            dplyr::select(gene_id, EA_signif),
        by = "gene_id"
    ) %>%
    dplyr::filter(BE_signif | EA_signif) %>%
    dplyr::mutate(
        present = dplyr::case_when(
            BE_signif & EA_signif ~ "BE+EA",
            BE_signif ~ "BE",
            EA_signif ~ "EA"
        ),
        present = factor(present, levels = c("BE", "EA", "BE+EA"))
    ) %>%
    dplyr::left_join(
        biotypes %>% dplyr::select(gene_id, gene_biotype),
        by = "gene_id"
    ) %>%
    dplyr::mutate(gene_biotype = factor(gene_biotype, levels = names(barretts_palette))) %>%
    dplyr::bind_rows(expand.grid(
        gene_biotype = unique(.$gene_biotype),
        present = factor(c("BE", "EA", "BE+EA"), levels = c("BE", "EA", "BE+EA"))
    )) %>%
    dplyr::summarize(
        .by = c("gene_biotype", "present"),
        count = dplyr::n() - 1
    ) %>%
    ggplot2::ggplot(ggplot2::aes(x = present, y = count, fill = gene_biotype)) +
    ggplot2::geom_bar(stat = "identity", position = 'dodge') +
    ggupset::axis_combmatrix(sep="\\+") +
    ggplot2::scale_y_continuous(expand = c(0,0)) +
    ggplot2::scale_fill_manual(values = barretts_palette) +
    ggplot2::labs(
        x = "Differentially expressed genes",
        y = "Number of genes",
        fill = "Gene biotype",
        title = "DE genes enriched in healthy controls"
    ) +
    theme_cell() +
    ggplot2::theme(legend.position = "none")
ggplot2::ggsave(
    filename = "results/panels/figure2_panelD.pdf",
    plot = fig2.panelD,
    width = 4, height = 4,
    device = cairo_pdf
)
ggplot2::ggsave(
    filename = "results/panels/figure2_panelD.png",
    plot = fig2.panelD,
    width = 4, height = 4, dpi = 600
)
```

## Panel E: Mitochondrial detection
```{r fig2.panel_e}
fig2.panelE <- DESeq2::counts(novel_dds, normalized = TRUE) %>%
    as.data.frame() %>%
    dplyr::mutate(gene_id = rownames(.)) %>%
    dplyr::left_join(
        biotypes %>%
        dplyr::select(gene_id, gene_biotype),
        by = "gene_id"
    ) %>%
    dplyr::filter(gene_biotype == "Mitochondrial") %>%
    tidyr::pivot_longer(
        cols = dplyr::starts_with("SLDK"),
        names_to = "names",
        values_to = "count"
    ) %>%
    dplyr::left_join(
        SummarizedExperiment::colData(novel_se) %>%
        as.data.frame(),
        by = "names"
    ) %>%
    dplyr::mutate(
        condition_code = factor(
            condition_code,
            levels = c("CT", "BE", "EA")
        )
    ) %>%
    dplyr::summarize(
        .by = c("gene_id", "condition_code"),
        count = mean(count)
    ) %>%
    ggplot2::ggplot(ggplot2::aes(x = condition_code, y = count, fill = condition_code)) +
    ggplot2::geom_boxplot(outlier.size = 0.5) +
    ggplot2::scale_y_log10(
        breaks = c(1e2, 1e5, 1e8),
        labels = c(
            expression(10^2),
            expression(10^5),
            expression(10^8)
        )
    ) +
    ggplot2::scale_fill_manual(values = c(
        "CT" = "#666666",
        "BE" = "#2C9DEF",
        "EA" = "#D95F02"
    )) +
    ggplot2::labs(
        x = "Sample Condition",
        y = "Normalized gene expression",
        title = "Mitochondrial gene expression"
    ) +
    ggsignif::geom_signif(
        comparisons = list(
            c("CT", "BE"),
            c("CT", "EA"),
            c("BE", "EA")
        ),
        map_signif_level = TRUE,
        y_position = c(6.5, 7.5, 6.75),
        textsize = 3, tip_length = 0,
        test = 'wilcox.test', test.args = list(paired = TRUE)
    ) +
    theme_cell() +
    ggplot2::theme(legend.position = "none") +
    ggplot2::coord_cartesian(clip = 'off')
ggplot2::ggsave(
    filename = "results/panels/figure2_panelE.pdf",
    plot = fig2.panelE,
    width = 3, height = 4,
    device = cairo_pdf
)
ggplot2::ggsave(
    filename = "results/panels/figure2_panelE.png",
    plot = fig2.panelE,
    width = 3, height = 4, dpi = 600
)
```

## Panel F: Mitochondrial heatmap
```{r fig2.panel_f}
## Orders dendrogram left to right by mean expression
treeorder <- function(hclust, values) {
    reorder_node <- function(node, hclust, values) {
        if (node < 0) return(node)

        left_child <- hclust$merge[node, 1]
        left_child <- reorder_node(left_child, hclust, values)
        left_leaves <- get_leaves(left_child, hclust)
        left_mean <- mean(values[left_leaves])

        right_child <- hclust$merge[node, 2]
        right_child <- reorder_node(right_child, hclust, values)
        right_leaves <- get_leaves(right_child, hclust)
        right_mean <- mean(values[right_leaves])

        if (left_mean > right_mean) {
            hclust$merge[node, 1] <<- right_child
            hclust$merge[node, 2] <<- left_child
        } else {
            hclust$merge[node, 1] <<- left_child
            hclust$merge[node, 2] <<- right_child
        }

        return(node)
    }

    get_leaves <- function(node, hclust) {
        if (node < 0) return(-node)

        c(
            get_leaves(hclust$merge[node, 1], hclust),
            get_leaves(hclust$merge[node, 2], hclust)
        )
    }

    reorder_node(nrow(hclust$merge), hclust, values)
    hclust$order <- get_leaves(nrow(hclust$merge), hclust)
    return(hclust)
}

mito_matrix <- DESeq2::counts(novel_dds, normalized = TRUE) %>%
    as.data.frame() %>%
    dplyr::mutate(gene_id = rownames(.)) %>%
    dplyr::left_join(
        biotypes,
        by = "gene_id"
    ) %>%
    dplyr::filter(gene_biotype == "Mitochondrial") %>%
    magrittr::set_rownames(.$gene_name) %>%
    dplyr::select(-gene_biotype, -gene_id, -gene_name) %>%
    magrittr::add(1) %>%
    log10() %>%
    sweep(1, rowMeans(.), '-') %>%
    sweep(1, apply(., 1, sd), '/')

mito_row_clusters <- mito_matrix %>%
    dist() %>%
    hclust() %>%
    treeorder(rowMeans(mito_matrix))
mito_row_dendro <- as.dendrogram(mito_row_clusters) %>%
    ggdendro::dendro_data()
mito_col_clusters <- mito_matrix %>%
    t() %>%
    dist() %>%
    hclust() %>%
    treeorder(colMeans(mito_matrix))
mito_col_dendro <- as.dendrogram(mito_col_clusters) %>%
    ggdendro::dendro_data()

mito_heatmap_data <- mito_matrix[mito_row_clusters$order, mito_col_clusters$order] %>%
    dplyr::mutate(Row = rownames(.)) %>%
    tidyr::pivot_longer(-Row, names_to = "Col", values_to = "Value") %>%
    dplyr::mutate(
        Row = factor(Row, levels = rownames(mito_matrix[mito_row_clusters$order,])),
        Col = factor(Col, levels = colnames(mito_matrix[,mito_col_clusters$order]))
    )

mito_heatmap <- mito_heatmap_data %>%
    ggplot2::ggplot(ggplot2::aes(x=Col, y=Row, fill=Value)) +
    ggplot2::geom_tile() +
    ggplot2::scale_fill_gradient2(
        low = RColorBrewer::brewer.pal(n = 11, name = "PRGn")[1],
        mid = "white",
        high = RColorBrewer::brewer.pal(n = 11, name = "PRGn")[11],
        limits = c(-max(abs(mito_matrix)), max(abs(mito_matrix))),
        name = "Z-Expression"
    ) +
    theme_cell() +
    ggplot2::theme(
        panel.grid = ggplot2::element_blank(),
        axis.title = ggplot2::element_blank(),
        axis.line = ggplot2::element_blank(),
        axis.ticks = ggplot2::element_blank(),
        axis.text.x = ggplot2::element_blank(),
        axis.text.y = ggplot2::element_text(size = 6, hjust = 0),
        plot.margin = ggplot2::margin(t = -0.1, r = -0.1, b = -0.1, l = -0.1),
        plot.tag = ggplot2::element_blank()
    )

mito_heatmap_col_annotations <- metadata %>%
    dplyr::filter(sample %in% colnames(mito_matrix)) %>%
    dplyr::select(sample, condition_code) %>%
    dplyr::mutate(sample = factor(sample, levels = levels(mito_heatmap_data$Col))) %>%
    ggplot2::ggplot(ggplot2::aes(x = sample, y = 1, fill = condition_code)) +
    ggplot2::geom_tile() +
    ggplot2::scale_fill_manual(values = c(
        "CT" = "#666666",
        "BE" = "#2C9DEF",
        "EA" = "#D95F02"
    )) +
    ggplot2::labs(fill = "Condition") +
    theme_cell() +
    ggplot2::theme(
        axis.line.x = ggplot2::element_blank(),
        axis.line.y = ggplot2::element_blank(),
        axis.title.x = ggplot2::element_blank(),
        axis.title.y = ggplot2::element_blank(),
        axis.ticks.x = ggplot2::element_blank(),
        axis.ticks.y = ggplot2::element_blank(),
        axis.text.x = ggplot2::element_blank(),
        axis.text.y = ggplot2::element_blank(),
        panel.grid = ggplot2::element_blank(),
        plot.margin = ggplot2::margin(t = -0.1, r = -0.1, b = -0.1, l = -0.1),
        plot.tag = ggplot2::element_blank(),
        legend.position = 'right'
    )

mito_heatmap_row_dendro <- ggdendro::segment(mito_row_dendro) %>%
    ggplot2::ggplot(ggplot2::aes(x = y, y = x, xend = yend, yend = xend)) +
    ggplot2::geom_segment() +
    ggplot2::scale_y_continuous(expand = c(0,0.5)) +
    ggplot2::scale_x_continuous(transform = 'reverse') +
    theme_cell() +
    ggdendro::theme_dendro() +
    ggplot2::theme(
        plot.margin = ggplot2::margin(t = -0.1, r = -0.1, b = -0.1, l = -0.1),
        plot.tag = ggplot2::element_blank()
    ) +
    ggplot2::labs(x = NULL, y = NULL)

mito_heatmap_col_dendro <- ggdendro::segment(mito_col_dendro) %>%
    ggplot2::ggplot(ggplot2::aes(x = x, y = y, xend = xend, yend = yend)) +
    ggplot2::geom_segment() +
    ggplot2::scale_x_continuous(expand = c(0,0.5)) +
    theme_cell() +
    ggdendro::theme_dendro() +
    ggplot2::theme(
        plot.margin = ggplot2::margin(t = -0.1, r = -0.1, b = -0.1, l = -0.1)
    ) +
    ggplot2::labs(x = NULL, y = NULL, title = "Mitochondrial expression heatmap") +
    ggplot2::coord_cartesian(clip = 'off')

fig2.panelF <- patchwork::wrap_plots(
        list(
            mito_heatmap_col_dendro, mito_heatmap_col_annotations,
            mito_heatmap_row_dendro,
            mito_heatmap
        ),
        design = '#A\n#B\nCD',
        widths = c(0.5, 4.235),
        heights = c(0.5, 0.235, 4)
    ) +
    patchwork::plot_layout(guides = 'collect')
ggplot2::ggsave(
    filename = "results/panels/figure2_panelF.pdf",
    plot = fig2.panelF,
    width = 17.4, height = 8.7, units = "cm",
    device = cairo_pdf
)
ggplot2::ggsave(
    filename = "results/panels/figure2_panelF.png",
    plot = fig2.panelF,
    width = 17.4, height = 8.7, units = "cm",
    dpi = 600
)
```

## Figure 2
```{r figure2}
figure2 <- patchwork::wrap_plots(
        list(
            fig2.panelAB[[1]][[1]], fig2.panelAB[[2]][[1]],
            fig2.panelC,
            fig2.panelD,
            fig2.panelE,
            mito_heatmap_col_dendro, mito_heatmap_col_annotations,
            mito_heatmap_row_dendro,
            mito_heatmap
        ),
        guides = 'collect',
        tag_level = 'keep',
        design = "AAAB\nCCCD\nE#FF\nE#GG\nEHII",
        heights = c(3.1567, 3.1567, .5, .235, 4),
        widths = c(
            2.3675,
            0.118375,
            0.68375, 3.55125
        )
    ) +
    patchwork::plot_annotation(tag_levels = 'A')
ggplot2::ggsave(
    filename = "results/figures/figure2.pdf",
    plot = figure2,
    width = 17.4, height = 13.5, units = "cm",
    device = cairo_pdf
)
ggplot2::ggsave(
    filename = "results/figures/figure2.png",
    plot = figure2,
    width = 17.4, height = 13.5, units = "cm",
    dpi = 600
)
```



# Figure 3 - Novel isoforms

## Panel A: Novel isoform sharing
```{r fig3.panel_a}
fig3.panelA <- novel_se %>%
    SummarizedExperiment::assays() %>%
    magrittr::extract2("counts") %>%
    { . > 1 } %>%
    rowSums() %>%
    data.frame() %>%
    magrittr::set_colnames('detected') %>%
    dplyr::mutate(gene_id = rownames(.)) %>%
    dplyr::filter(startsWith(gene_id, 'novel_gene')) %>%
    ggplot2::ggplot(ggplot2::aes(x = detected)) +
    ggplot2::geom_bar(fill = barretts_palette["Novel Gene"]) +
    ggplot2::geom_vline(
        xintercept = 21,
        color = "#282A36",
        linetype = "dashed"
    ) +
    ggplot2::labs(
        x = "Number of samples with expression",
        y = "Number of novel isoforms",
        title = "Common expression of novel cfRNAs"
    ) +
    theme_cell() +
    ggplot2::scale_y_continuous(expand = c(0,0))
ggplot2::ggsave(
    filename = "results/panels/figure3_panelA.pdf",
    plot = fig3.panelA,
    width = 5.8, height = 5.8, units = "cm",
    device = cairo_pdf
)
ggplot2::ggsave(
    filename = "results/panels/figure3_panelA.png",
    plot = fig3.panelA,
    width = 5.8, height = 5.8, units = "cm",
    dpi = 600
)
```

## Panel B: Unique novel isoform detection
```{r fig3.panel_b}
fig3.panelB <- novel_se %>%
    SummarizedExperiment::assays() %>%
    magrittr::extract2("counts") %>%
    { . > 1 } %>%
    as.data.frame() %>%
    dplyr::filter(startsWith(rownames(.), 'novel_gene_')) %>%
    dplyr::filter(rowSums(.) == 1) %>%
    dplyr::mutate(gene_id = rownames(.)) %>%
    tidyr::pivot_longer(
        cols = dplyr::starts_with("SLDK"),
        names_to = "sample",
        values_to = "detected"
    ) %>%
    dplyr::filter(detected) %>%
    dplyr::left_join(
        metadata %>%
        dplyr::select(sample, condition_code),
        by = "sample"
    ) %>%
    dplyr::summarize(
        .by = c("condition_code", "sample"),
        detected = dplyr::n()
    ) %>%
    ggplot2::ggplot(ggplot2::aes(x = condition_code, y = detected, fill = condition_code)) +
    ggplot2::geom_boxplot(outlier.size = 0.5) +
    ggplot2::scale_fill_manual(values = c(
        "CT" = "#666666",
        "BE" = "#2C9DEF",
        "EA" = "#D95F02"
    )) +
    ggplot2::labs(
        x = "Sample Condition",
        y = "Number of unique novel isoforms",
        title = "Patient-specific novel cfRNAs"
    ) +
    ggsignif::geom_signif(
        comparisons = list(
            c("CT", "BE"),
            c("BE", "EA"),
            c("CT", "EA")
        ),
        map_signif_level = TRUE,
        y_position = c(660, 720, 780),
        textsize = 2, tip_length = 0,
        test = 'wilcox.test'
    ) +
    theme_cell() +
    ggplot2::theme(legend.position = "none") +
    ggplot2::coord_cartesian(clip = 'off')
ggplot2::ggsave(
    filename = "results/panels/figure3_panelB.pdf",
    plot = fig3.panelB,
    width = 5.8, height = 6.5, units = "cm",
    device = cairo_pdf
)
ggplot2::ggsave(
    filename = "results/panels/figure3_panelB.png",
    plot = fig3.panelB,
    width = 5.8, height = 6.5, units = "cm",
    dpi = 600
)
```

## Panel C: Novel isoform heatmap
```{r fig3.panel_c}
novel_de <- DESeq2::results(novel_dds, contrast = c("condition_code", "EA", "CT")) %>%
    as.data.frame() %>%
    dplyr::filter(!is.na(padj)) %>%
    dplyr::mutate(
        gene_id = rownames(.),
        signif = (padj < pval_thresh) & (abs(log2FoldChange) >= logFC_thresh)
    ) %>%
    dplyr::filter(signif) %>%
    dplyr::mutate(enriched = dplyr::case_when(
            log2FoldChange > 0 ~ "EA-enriched",
            log2FoldChange < 0 ~ "CT-enriched"
        )) %>%
    dplyr::left_join(
        biotypes,
        by = "gene_id"
    ) %>%
    dplyr::filter(gene_biotype == "Novel Gene") %>%
    dplyr::bind_rows(
        DESeq2::results(novel_dds, contrast = c("condition_code", "BE", "CT")) %>%
        as.data.frame() %>%
        dplyr::filter(!is.na(padj)) %>%
        dplyr::mutate(
            gene_id = rownames(.),
            signif = (padj < pval_thresh) & (abs(log2FoldChange) >= logFC_thresh),
        ) %>%
        dplyr::filter(signif) %>%
        dplyr::mutate(enriched = dplyr::case_when(
            log2FoldChange > 0 ~ "BE-enriched",
            log2FoldChange < 0 ~ "CT-enriched"
        )) %>%
        dplyr::left_join(
            biotypes,
            by = "gene_id"
        ) %>%
        dplyr::filter(gene_biotype == "Novel Gene")
    )

novel_matrix <- DESeq2::counts(novel_dds, normalized = TRUE) %>%
    as.data.frame() %>%
    dplyr::filter(rownames(.) %in% novel_de$gene_id) %>%
    dplyr::mutate(gene_id = rownames(.)) %>%
    dplyr::left_join(
        biotypes,
        by = "gene_id"
    ) %>%
    dplyr::filter(gene_biotype == "Novel Gene") %>%
    magrittr::set_rownames(.$gene_name) %>%
    dplyr::select(-gene_biotype, -gene_id, -gene_name) %>%
    magrittr::add(1) %>%
    log10() %>%
    sweep(1, rowMeans(.), '-') %>%
    sweep(1, apply(., 1, sd), '/')

novel_col_clusters <- novel_matrix %>%
    t() %>%
    dist() %>%
    hclust() %>%
    treeorder(colMeans(novel_matrix))
novel_col_dendro <- as.dendrogram(novel_col_clusters) %>%
    ggdendro::dendro_data()

novel_row_clusters <- novel_matrix %>%
    dist() %>%
    hclust() %>%
    treeorder(
        novel_de %>%
        {
            dat <- .
            res <- dat$log2FoldChange
            names(res) <- dat$gene_name
            res
        }
    )
novel_row_dendro <- as.dendrogram(novel_row_clusters) %>%
    ggdendro::dendro_data()

novel_heatmap_data <- novel_matrix[novel_row_clusters$order, novel_col_clusters$order] %>%
    dplyr::mutate(Row = rownames(.)) %>%
    tidyr::pivot_longer(-Row, names_to = "Col", values_to = "Value") %>%
    dplyr::mutate(
        Row = factor(Row, levels = rownames(novel_matrix[novel_row_clusters$order,])),
        Col = factor(Col, levels = colnames(novel_matrix[,novel_col_clusters$order]))
    )

novel_heatmap <- novel_heatmap_data %>%
    ggplot2::ggplot(ggplot2::aes(x=Col, y=Row, fill=Value)) +
    ggplot2::geom_tile() +
    ggplot2::scale_fill_gradient2(
        low = RColorBrewer::brewer.pal(n = 11, name = "PRGn")[1],
        mid = "white",
        high = RColorBrewer::brewer.pal(n = 11, name = "PRGn")[11],
        limits = c(-max(abs(novel_matrix)), max(abs(novel_matrix))),
        name = "Z-Expression"
    ) +
    theme_cell() +
    ggplot2::theme(
        panel.grid = ggplot2::element_blank(),
        panel.spacing = ggplot2::unit(0, "pt"),
        axis.title = ggplot2::element_blank(),
        axis.line = ggplot2::element_blank(),
        axis.ticks = ggplot2::element_blank(),
        axis.text.x = ggplot2::element_blank(),
        axis.text.y = ggplot2::element_blank(),
        plot.tag = ggplot2::element_blank(),
        plot.margin = ggplot2::margin(t = -0.1, r = -0.1, b = -0.1, l = -0.1)
    )

novel_heatmap_col_annotations <- metadata %>%
    dplyr::filter(sample %in% colnames(novel_matrix)) %>%
    dplyr::select(sample, condition_code) %>%
    dplyr::mutate(sample = factor(sample, levels = levels(novel_heatmap_data$Col))) %>%
    ggplot2::ggplot(ggplot2::aes(x = sample, y = 1, fill = condition_code)) +
    ggplot2::geom_tile() +
    ggplot2::scale_fill_manual(values = c(
        "CT" = "#666666",
        "BE" = "#2C9DEF",
        "EA" = "#D95F02"
    )) +
    ggplot2::labs(fill = "Sample Condition") +
    theme_cell() +
    ggplot2::theme(
        axis.line.x = ggplot2::element_blank(),
        axis.line.y = ggplot2::element_blank(),
        axis.title.x = ggplot2::element_blank(),
        axis.title.y = ggplot2::element_blank(),
        axis.ticks.x = ggplot2::element_blank(),
        axis.ticks.y = ggplot2::element_blank(),
        axis.text.x = ggplot2::element_blank(),
        axis.text.y = ggplot2::element_blank(),
        panel.grid = ggplot2::element_blank(),
        panel.spacing = ggplot2::unit(0, "pt"),
        plot.tag = ggplot2::element_blank(),
        plot.margin = ggplot2::margin(t = -0.1, r = -0.1, b = -0.1, l = -0.1),
        legend.position = 'right'
    )

novel_heatmap_row_annotations <- novel_de %>%
    dplyr::select(gene_name, enriched) %>%
    dplyr::summarize(
        .by = gene_name,
        enriched = paste(sort(unique(enriched)), collapse = " & ") %>%
            stringr::str_replace_all('-enriched', '') %>%
            paste0('-enriched')
    ) %>%
    dplyr::filter(gene_name %in% rownames(novel_matrix)) %>%
    dplyr::select(gene_name, enriched) %>%
    dplyr::mutate(enriched = factor(enriched, levels = c("CT-enriched", "BE-enriched", "EA-enriched", "BE & EA-enriched"))) %>%
    dplyr::mutate(gene_name = factor(gene_name, levels = levels(novel_heatmap_data$Row))) %>%
    ggplot2::ggplot(ggplot2::aes(x = 1, y = gene_name, fill = enriched)) +
    ggplot2::geom_tile() +
    ggplot2::scale_fill_manual(values = c(
        "CT-enriched" = "#666666",
        "BE-enriched" = "#2C9DEF",
        "EA-enriched" = "#D95F02",
        "BE & EA-enriched" = "#A6761D"
    )) +
    ggplot2::labs(fill = "Gene Enrichment") +
    theme_cell() +
    ggplot2::theme(
        axis.line.x = ggplot2::element_blank(),
        axis.line.y = ggplot2::element_blank(),
        axis.title.x = ggplot2::element_blank(),
        axis.title.y = ggplot2::element_blank(),
        axis.ticks.x = ggplot2::element_blank(),
        axis.ticks.y = ggplot2::element_blank(),
        axis.text.x = ggplot2::element_blank(),
        axis.text.y = ggplot2::element_blank(),
        panel.grid = ggplot2::element_blank(),
        panel.spacing = ggplot2::unit(0, "pt"),
        plot.tag = ggplot2::element_blank(),
        plot.margin = ggplot2::margin(t = -0.1, r = -0.1, b = -0.1, l = -0.1),
        legend.position = 'right'
    )

novel_heatmap_row_dendro <- ggdendro::segment(novel_row_dendro) %>%
    ggplot2::ggplot(ggplot2::aes(x = y, y = x, xend = yend, yend = xend)) +
    ggplot2::geom_segment() +
    ggplot2::scale_y_continuous(expand = c(0,0.5)) +
    ggplot2::scale_x_continuous(transform = 'reverse') +
    theme_cell() +
    ggdendro::theme_dendro() +
    ggplot2::theme(
        plot.tag = ggplot2::element_blank(),
        plot.margin = ggplot2::margin(t = -0.1, r = -0.1, b = -0.1, l = -0.1)
    ) +
    ggplot2::labs(x = NULL, y = NULL)

novel_heatmap_col_dendro <- ggdendro::segment(novel_col_dendro) %>%
    ggplot2::ggplot(ggplot2::aes(x = x, y = y, xend = xend, yend = yend)) +
    ggplot2::geom_segment() +
    ggplot2::scale_x_continuous(expand = c(0,0.5)) +
    theme_cell() +
    ggdendro::theme_dendro() +
    ggplot2::theme(
        plot.tag = ggplot2::element_blank(),
        plot.margin = ggplot2::margin(t = -0.1, r = -0.1, b = -0.1, l = -0.1)
    ) +
    ggplot2::labs(x = NULL, y = NULL)

spacer <- ggplot2::ggplot() +
    theme_cell() +
    ggplot2::theme_void() +
    ggplot2::theme(
        plot.title = ggplot2::element_text(
            size = 6,
            color = "#282A36",
            face = 'bold',
            hjust = 0,
            vjust = 0.5
        ),
        plot.tag = ggplot2::element_text(
            size = 10,
            face = "bold",
            color = "#282A36"
        ),
    ) +
    ggplot2::labs(title = "DE novel cfRNA expression heatmap") +
    ggplot2::coord_cartesian(clip = 'off')

fig3.panelC <- patchwork::wrap_plots(
        list(
            spacer,
            novel_heatmap_col_dendro,
            novel_heatmap_col_annotations,
            novel_heatmap_row_dendro,
            novel_heatmap_row_annotations,
            novel_heatmap
        ),
        design = 'AAB\nAAC\nDEF',
        widths = c(0.5, 0.088, 4),
        heights = c(0.5, 0.036, 4)
    ) +
    patchwork::plot_layout(guides = 'collect')
ggplot2::ggsave(
    filename = "results/panels/figure3_panelC.pdf",
    plot = fig3.panelC,
    width = 17.4, height = 8.7, units= "cm",
    device = cairo_pdf
)
ggplot2::ggsave(
    filename = "results/panels/figure3_panelC.png",
    plot = fig3.panelC,
    width = 17.4, height = 8.7, units= "cm",
    dpi = 600
)
```

# Figure 3
```{r figure3}
figure3 <- patchwork::wrap_plots(
        list(
            fig3.panelA,
            fig3.panelB,
            spacer,
            novel_heatmap_col_dendro,
            novel_heatmap_col_annotations,
            novel_heatmap_row_dendro,
            novel_heatmap_row_annotations,
            novel_heatmap
        ),
        guides = 'collect',
        tag_level = 'keep',
        design = "AAAB\nCCDD\nCCEE\nFGHH",
        heights = c(3.024, 0.45, 0.086, 4),
        widths = c(0.5, 0.088, 2.853, 1.147)
    ) +
    patchwork::plot_annotation(tag_levels = 'A')
ggplot2::ggsave(
    filename = "results/figures/figure3.pdf",
    plot = figure3,
    width = 17.4, height = 11.6, units = "cm",
    device = cairo_pdf
)
ggplot2::ggsave(
    filename = "results/figures/figure3.png",
    plot = figure3,
    width = 17.4, height = 11.6, units = "cm",
    dpi = 600
)
```



# Figure 4: Modeling disease vs. controls
```{r modeling_helper}
glm_optimization <- function(
    y, X, n_repeats = 10, n_folds = 4,
    final_folds = caret::createFolds(y, k = 4, returnTrain = TRUE)
) {
    ## Train LASSO models across lambda grid
    model_list <- foreach::foreach(1:n_repeats, .combine = 'c') %do% {
        weak_models <- glmnet::cv.glmnet(
            x = X, y = y,
            alpha = 1, lambda = seq(0.01, 1, by = 0.01),
            family = 'binomial', type.measure = 'deviance',
            nfolds = n_folds
        )

        return(list(list(
            lambda_min = weak_models$lambda.min,
            lambda_1se = weak_models$lambda.1se,
            mse = weak_models$cvm[weak_models$index[1]],
            model = weak_models
        )))
    }

    ## Find best feature set across repeats
    best_lambda_model <- model_list[[which.min(unlist(lapply(
        model_list,
        function(x) { x$mse }
    )))]]

    best_features <- best_lambda_model$model$glmnet.fit$beta[,best_lambda_model$model$index[1]] %>%
        { . != 0 } %>%
        unname()

    if (sum(best_features) < 2) {
        cat("Too many features removed with LASSO; using all features.\n")
        best_features <- rep(TRUE, ncol(X))
    }

    alpha_models <- foreach::foreach(alpha = seq(0, 1, by = 0.01), .combine = 'c') %do% {
        X_best <- X[,best_features] %>%
            as.matrix() %>%
            magrittr::set_colnames(colnames(X)[best_features])

        alpha_model <- glmnet::cv.glmnet(
            x = X_best, y = y,
            alpha = alpha,
            lambda = seq(0.01, 1, by = 0.01),
            family = 'binomial',
            type.measure = 'deviance',
            nfolds = n_folds
        )

        return(list(list(
            alpha = alpha,
            lambda_min = alpha_model$lambda.min,
            lambda_1se = alpha_model$lambda.1se,
            model = alpha_model,
            mse = alpha_model$cvm[alpha_model$index[1]]
        )))
    }

    ## Get final best model
    tuned_model <- alpha_models[[which.min(unlist(lapply(
        alpha_models,
        function(x) { x$mse }
    )))]]

    final_train <- setdiff(1:nrow(X_best), final_fold)

    final_model <- glmnet::glmnet(
        x = X_best[final_train,], y = y[final_train],
        alpha = tuned_model$alpha,
        lambda = tuned_model$lambda_min,
        family = 'binomial'
    )

    return(list(
        final_model = final_model,
        best_features = colnames(X)[best_features],
        mse = tuned_model$mse,
        alpha = tuned_model$alpha,
        lambda = tuned_model$lambda_min,
        training_idx = final_train,
        testing_idx = final_fold
    ))
}
```

## Panel A-D: Barretts modeling
```{r fig4.panel_abcd}
de_genes_be <- DESeq2::results(
        novel_dds,
        contrast = c("condition_code", "BE", "CT")
    ) %>%
    as.data.frame() %>%
    dplyr::mutate(gene_id = rownames(.)) %>%
    dplyr::left_join(SummarizedExperiment::rowData(novel_se) %>%
        as.data.frame() %>%
        dplyr::select(gene_id, gene_name, gene_biotype),
        by = "gene_id"
    ) %>%
    dplyr::mutate(
        condition = dplyr::case_when(
            log2FoldChange > 1 & padj < 0.05 ~ "BE-enriched",
            log2FoldChange < -1 & padj < 0.05 ~ "CT-enriched",
            TRUE ~ "Non-DE"
        )
    )
if (!file.exists('results/models/models_be.rds')) {
    modeling_metadata_be <- SummarizedExperiment::colData(novel_dds) %>%
        as.data.frame() %>%
        dplyr::filter(condition != "Esophageal Adenocarcinoma") %>%
        dplyr::mutate(feature_sex = dplyr::case_when(
            sex == 'F' ~ 0,
            TRUE ~ 1
        )) %>%
        dplyr::select(feature_sex, scaled_age, condition) %>%
        dplyr::mutate(
            condition = factor(condition, levels = c("Control", "Barrett's Esophagus"))
        )

    modeling_counts_be <- novel_dds %>%
        DESeq2::varianceStabilizingTransformation() %>%
        SummarizedExperiment::assay() %>%
        as.data.frame() %>%
        dplyr::select(dplyr::all_of(rownames(modeling_metadata_be))) %>%
        dplyr::filter(rownames(.) %in% de_genes_be[de_genes_be$condition != "Non-DE",]$gene_id) %>%
        t() %>%
        as.data.frame()

    y_be = modeling_metadata_be %>%
        dplyr::pull(condition)

    X_metadata_be <- modeling_metadata_be %>%
        dplyr::select(feature_sex, scaled_age) %>%
        as.matrix()

    X_counts_be <- modeling_counts_be %>%
        as.matrix()

    X_gencode_be <- modeling_counts_be %>%
        as.data.frame() %>%
        dplyr::select(dplyr::starts_with("ENSG")) %>%
        as.matrix()

    X_novel_be <- modeling_counts_be %>%
        as.data.frame() %>%
        dplyr::select(dplyr::starts_with("novel_gene_")) %>%
        as.matrix()

    set.seed(1337)
    final_fold <- caret::createFolds(y_be, k = 3)$Fold1

    models_be <- sapply(
        list(
            metadata = X_metadata_be,
            gencode = X_gencode_be,
            novel = X_novel_be,
            counts = X_counts_be
        ),
        function(X) {
            cat("Training model...\n")
            model <- glm_optimization(y_be, X, final_folds = final_fold)
            cat("Collecting results...\n")
            features <- coef(model$final_model, s = "lambda.min") %>%
                as.matrix() %>%
                as.data.frame() %>%
                dplyr::mutate(feature = rownames(.)) %>%
                dplyr::filter(s1 != 0)
            pred <- predict(
                model$final_model,
                newx = X[model$testing_idx, model$best_features],
                s = "lambda.min", type = "response"
            )
            roc <- pROC::roc(y_be[model$testing_idx], pred[,1], quiet=TRUE)
            cat("...Done!\n")
            return(list(
                model = model,
                features = features,
                roc = roc
            ))
        }
    )
    readr::write_rds(models_be, 'results/models/models_be.rds')
} else {
    models_be <- readr::read_rds('results/models/models_be.rds')
}

figure4.panelABCD <- lapply(c('metadata', 'gencode', 'novel', 'counts'), function(X) {
    titles <- c(
        'metadata' = 'Metadata',
        'gencode' = 'GENCODE',
        'novel' = 'Novel cfRNA',
        'counts' = 'GENCODE+novel cfRNA'
    )
    plot <- models_be['roc',X] %>%
        magrittr::set_names(X) %>%
        pROC::ggroc(aes = c("color")) +
        theme_cell() +
        ggplot2::geom_abline(
            slope = 1, intercept = 1,
            linetype = 'dashed', color = 'lightgray'
        ) +
        ggplot2::geom_text(
            x = 0, y = 0, hjust = 1, vjust = 0, color = 'black',
            label = paste0(
                "AUC = ",
                round(as.numeric(pROC::auc(models_be['roc', X][[1]])), digits=2),
                "\n",
                nrow(models_be['features', X][[1]])-1,
                " non-zero features"
            ),
            check_overlap = TRUE
        ) +
        ggplot2::scale_color_manual(values = c(
        "metadata" = "#666666",
        "gencode" = "#E6AB02",
        "novel" = "#2C9DEF",
        "counts" = "#1B9E77"
        )) +
        ggplot2::labs(
            x = "False Positive Rate",
            y = "True Positive Rate"
        ) +
        ggplot2::theme(legend.position = "none")
    lines <- which(lapply(plot$layers, function(l) {class(l$geom)[1]}) == "GeomLine")
    plot$layers = c(plot$layers[-lines], plot$layers[lines])
    panel = c('E', 'F', 'G', 'H')[which(c('metadata', 'gencode', 'novel', 'counts') == X)]
    ggplot2::ggsave(
        filename = paste0("results/panels/figure4_panel", panel, ".pdf"),
        plot = plot,
        width = 4, height = 4,
        device = cairo_pdf
    )
    ggplot2::ggsave(
        filename = paste0("results/panels/figure4_panel", panel, ".png"),
        plot = plot,
        width = 4, height = 4, dpi = 600
    )
    return(list(plot))
})

# 5 of 13 count features are novel!
models_be[['features', 'counts']] %>%
    as.data.frame() %>%
    dplyr::left_join(biotypes, by = c("feature" = "gene_id")) %>%
    dplyr::arrange(desc(abs(s1)))
```

## Panel E-H: Esophagus modeling
```{r fig4.panel_efgh}
de_genes_ea <- DESeq2::results(
        novel_dds,
        contrast = c("condition_code", "EA", "CT")
    ) %>%
    as.data.frame() %>%
    dplyr::mutate(gene_id = rownames(.)) %>%
    dplyr::left_join(SummarizedExperiment::rowData(novel_se) %>%
        as.data.frame() %>%
        dplyr::select(gene_id, gene_name, gene_biotype),
        by = "gene_id"
    ) %>%
    dplyr::mutate(
        condition = dplyr::case_when(
            log2FoldChange > 1 & padj < 0.05 ~ "EA-enriched",
            log2FoldChange < -1 & padj < 0.05 ~ "CT-enriched",
            TRUE ~ "Non-DE"
        )
    )

if (!file.exists('results/models/models_ea.rds')) {
    modeling_metadata_ea <- SummarizedExperiment::colData(novel_dds) %>%
        as.data.frame() %>%
        dplyr::filter(condition != "Barrett's Esophagus") %>%
        dplyr::mutate(feature_sex = dplyr::case_when(
            sex == 'F' ~ 0,
            TRUE ~ 1
        )) %>%
        dplyr::select(feature_sex, scaled_age, condition) %>%
        dplyr::mutate(
            condition = factor(condition, levels = c("Control", "Esophageal Adenocarcinoma"))
        )

    modeling_counts_ea <- novel_dds %>%
        DESeq2::varianceStabilizingTransformation() %>%
        SummarizedExperiment::assay() %>%
        as.data.frame() %>%
        dplyr::select(dplyr::all_of(rownames(modeling_metadata_ea))) %>%
        dplyr::filter(rownames(.) %in% de_genes_ea[de_genes_ea$condition != "Non-DE",]$gene_id) %>%
        t() %>%
        as.data.frame()

    y_ea = modeling_metadata_ea %>%
        dplyr::pull(condition)

    X_metadata_ea <- modeling_metadata_ea %>%
        dplyr::select(feature_sex, scaled_age) %>%
        as.matrix()

    X_counts_ea <- modeling_counts_ea %>%
        as.matrix()

    X_gencode_ea <- modeling_counts_ea %>%
        as.data.frame() %>%
        dplyr::select(dplyr::starts_with("ENSG")) %>%
        as.matrix()

    X_novel_ea <- modeling_counts_ea %>%
        as.data.frame() %>%
        dplyr::select(dplyr::starts_with("novel_gene_")) %>%
        as.matrix()

    set.seed(1337)
    final_fold <- caret::createFolds(y_ea, k = 3)$Fold1

    models_ea <- sapply(
        list(
            metadata = X_metadata_ea,
            gencode = X_gencode_ea,
            novel = X_novel_ea,
            counts = X_counts_ea
        ),
        function(X) {
            cat("Training model...\n")
            model <- glm_optimization(y_ea, X, final_folds = final_fold)
            cat("Collecting results...\n")
            features <- coef(model$final_model, s = "lambda.min") %>%
                as.matrix() %>%
                as.data.frame() %>%
                dplyr::mutate(feature = rownames(.)) %>%
                dplyr::filter(s1 != 0)
            pred <- predict(
                model$final_model,
                newx = X[model$testing_idx, model$best_features],
                s = "lambda.min", type = "response"
            )
            roc <- pROC::roc(y_ea[model$testing_idx], pred[,1], quiet=TRUE)
            cat("...Done!\n")
            return(list(
                model = model,
                features = features,
                roc = roc
            ))
        }
    )
    readr::write_rds(models_ea, 'results/models/models_ea.rds')
} else {
    models_ea <- readr::read_rds('results/models/models_ea.rds')
}

figure4.panelEFGH <- lapply(c('metadata', 'gencode', 'novel', 'counts'), function(X) {
    titles <- c(
        'metadata' = 'Metadata',
        'gencode' = 'GENCODE',
        'novel' = 'Novel cfRNA',
        'counts' = 'GENCODE+novel cfRNA'
    )
    plot <- models_ea['roc',X] %>%
        magrittr::set_names(X) %>%
        pROC::ggroc(aes = c("color")) +
        theme_cell() +
        ggplot2::geom_abline(
            slope = 1, intercept = 1,
            linetype = 'dashed', color = 'lightgray'
        ) +
        ggplot2::geom_text(
            x = 0, y = 0, hjust = 1, vjust = 0, color = 'black',
            label = paste0(
                'AUC=',
                round(as.numeric(pROC::auc(models_ea['roc', X][[1]])), digits=2),
                "\n",
                nrow(models_ea['features', X][[1]])-1,
                " non-zero features"
            ),
            check_overlap = TRUE
        ) +
        ggplot2::scale_color_manual(values = c(
            "metadata" = "#666666",
            "gencode" = "#E6AB02",
            "novel" = "#2C9DEF",
            "counts" = "#1B9E77"
        )) +
        ggplot2::labs(
            x = "False Positive Rate",
            y = "True Positive Rate"
        ) +
        ggplot2::theme(legend.position = "none")
    lines <- which(lapply(plot$layers, function(l) {class(l$geom)[1]}) == "GeomLine")
    plot$layers = c(plot$layers[-lines], plot$layers[lines])
    panel = c('E', 'F', 'G', 'H')[which(c('metadata', 'gencode', 'novel', 'counts') == X)]
    ggplot2::ggsave(
        filename = paste0("results/panels/figure4_panel", panel, ".pdf"),
        plot = plot,
        width = 4, height = 4,
        device = cairo_pdf
    )
    ggplot2::ggsave(
        filename = paste0("results/panels/figure4_panel", panel, ".png"),
        plot = plot,
        width = 4, height = 4, dpi = 600
    )
    return(list(plot))
})

# 11 of 15 count features are novel!
models_ea[['features', 'counts']] %>%
    as.data.frame() %>%
    dplyr::left_join(biotypes, by = c("feature" = "gene_id")) %>%
    dplyr::arrange(desc(abs(s1)))
```

## Panel I-L: Disease modeling
```{r fig4.panel_ijkl}
de_genes_dis <- DESeq2::results(
        novel_dds,
        contrast = c("condition_code", "EA", "BE")
    ) %>%
    as.data.frame() %>%
    dplyr::mutate(gene_id = rownames(.)) %>%
    dplyr::left_join(SummarizedExperiment::rowData(novel_se) %>%
        as.data.frame() %>%
        dplyr::select(gene_id, gene_name, gene_biotype),
        by = "gene_id"
    ) %>%
    dplyr::mutate(
        condition = dplyr::case_when(
            log2FoldChange > 1 & padj < 0.05 ~ "EA-enriched",
            log2FoldChange < -1 & padj < 0.05 ~ "BE-enriched",
            TRUE ~ "Non-DE"
        )
    )
if (!file.exists('results/models/models_dis.rds')) {
    modeling_metadata_dis <- SummarizedExperiment::colData(novel_dds) %>%
        as.data.frame() %>%
        dplyr::filter(condition != "Control") %>%
        dplyr::mutate(feature_sex = dplyr::case_when(
            sex == 'F' ~ 0,
            TRUE ~ 1
        )) %>%
        dplyr::select(feature_sex, scaled_age, condition) %>%
        dplyr::mutate(
            condition = factor(condition, levels = c(
                "Barrett's Esophagus",
                "Esophageal Adenocarcinoma"
            ))
        )

    modeling_counts_dis <- novel_dds %>%
        DESeq2::varianceStabilizingTransformation() %>%
        SummarizedExperiment::assay() %>%
        as.data.frame() %>%
        dplyr::select(dplyr::all_of(rownames(modeling_metadata_dis))) %>%
        dplyr::filter(
            rownames(.) %in% de_genes_dis[de_genes_dis$condition != "Non-DE",]$gene_id
        ) %>%
        t() %>%
        as.data.frame()

    y_dis = modeling_metadata_dis %>%
        dplyr::pull(condition)

    X_metadata_dis <- modeling_metadata_dis %>%
        dplyr::select(feature_sex, scaled_age) %>%
        as.matrix()

    X_counts_dis <- modeling_counts_dis %>%
        as.matrix()

    X_gencode_dis <- modeling_counts_dis %>%
        as.data.frame() %>%
        dplyr::select(dplyr::starts_with("ENSG")) %>%
        as.matrix()

    X_novel_dis <- modeling_counts_dis %>%
        as.data.frame() %>%
        dplyr::select(dplyr::starts_with("novel_gene_")) %>%
        as.matrix()

    set.seed(1337)
    final_fold <- caret::createFolds(y_dis, k = 3)$Fold1

    models_dis <- sapply(
        list(
            metadata = X_metadata_dis,
            gencode = X_gencode_dis,
            novel = X_novel_dis,
            counts = X_counts_dis
        ),
        function(X) {
            cat("Training model...\n")
            model <- glm_optimization(y_dis, X, final_folds = final_fold)
            cat("Collecting results...\n")
            features <- coef(model$final_model, s = "lambda.min") %>%
                as.matrix() %>%
                as.data.frame() %>%
                dplyr::mutate(feature = rownames(.)) %>%
                dplyr::filter(s1 != 0)
            pred <- predict(
                model$final_model,
                newx = X[model$testing_idx, model$best_features],
                s = "lambda.min", type = "response"
            )
            roc <- pROC::roc(y_dis[model$testing_idx], pred[,1], quiet=TRUE)
            cat("...Done!\n")
            return(list(
                model = model,
                features = features,
                roc = roc
            ))
        }
    )
    readr::write_rds(models_dis, 'results/models/models_dis.rds')
} else {
    models_dis <- readr::read_rds('results/models/models_dis.rds')
}

figure4.panelIJKL <- lapply(c('metadata', 'gencode', 'novel', 'counts'), function(X) {
    titles <- c(
        'metadata' = 'Metadata',
        'gencode' = 'GENCODE',
        'novel' = 'Novel cfRNA',
        'counts' = 'GENCODE+novel cfRNA'
    )
    plot <- models_dis['roc',X] %>%
        magrittr::set_names(X) %>%
        pROC::ggroc(aes = c("color")) +
        theme_cell() +
        ggplot2::geom_abline(
            slope = 1, intercept = 1,
            linetype = 'dashed', color = 'lightgray'
        ) +
        ggplot2::geom_text(
            x = 0, y = 0, hjust = 1, vjust = 0, color = 'black',
            label = paste0(
                'AUC=',
                round(as.numeric(pROC::auc(models_dis['roc', X][[1]])), digits=2),
                "\n",
                nrow(models_dis['features', X][[1]])-1,
                " non-zero features"
            ),
            check_overlap = TRUE
        ) +
        ggplot2::scale_color_manual(values = c(
        "metadata" = "#666666",
        "gencode" = "#E6AB02",
        "novel" = "#2C9DEF",
        "counts" = "#1B9E77"
        )) +
        ggplot2::labs(
            x = "False Positive Rate",
            y = "True Positive Rate"
        ) +
        ggplot2::theme(legend.position = "none")
    lines <- which(lapply(plot$layers, function(l) {class(l$geom)[1]}) == "GeomLine")
    plot$layers = c(plot$layers[-lines], plot$layers[lines])
    panel = c('I', 'J', 'K', 'L')[which(c('metadata', 'gencode', 'novel', 'counts') == X)]
    ggplot2::ggsave(
        filename = paste0("results/panels/figure4_panel", panel, ".pdf"),
        plot = plot,
        width = 4, height = 4,
        device = cairo_pdf
    )
    ggplot2::ggsave(
        filename = paste0("results/panels/figure4_panel", panel, ".png"),
        plot = plot,
        width = 4, height = 4, dpi = 600
    )
    return(list(plot))
})

# 12 of 14 count features are novel!
models_dis[['features', 'counts']] %>%
    as.data.frame() %>%
    dplyr::left_join(biotypes, by = c("feature" = "gene_id")) %>%
    dplyr::arrange(desc(abs(s1)))
```

## Figure 4
```{r figure4}
plot_label <- function(title, rot = 0, size = 8) {
    lab <- ggplot2::ggplot() +
        ggplot2::theme_void() +
        ggplot2::theme(
            plot.title = ggplot2::element_text(
                size = size,
                color = "#282A36",
                face = 'bold',
                hjust = ifelse(rot == 0, 0.6, 0.5),
                vjust = ifelse(rot == 0, 0, 0.6),
                angle = rot
            ),
            panel.spacing = ggplot2::unit(0, "cm")
        ) +
        ggplot2::labs(title = title) +
        ggplot2::coord_cartesian(clip = 'off')
    return(patchwork::free(lab))
}
figure4 <- patchwork::wrap_plots(
        list(
            plot_label("Metadata"),
            plot_label("GENCODE"),
            plot_label("Novel cfRNA"),
            plot_label("GENCODE+Novel"),
            plot_label("Barrett's Esophagus\nfrom Healthy", 90, 6),
            figure4.panelABCD[[1]][[1]],
            figure4.panelABCD[[2]][[1]],
            figure4.panelABCD[[3]][[1]],
            figure4.panelABCD[[4]][[1]],
            plot_label("Esophageal Adenocarcinoma\nfrom Healthy", 90, 6),
            figure4.panelEFGH[[1]][[1]],
            figure4.panelEFGH[[2]][[1]],
            figure4.panelEFGH[[3]][[1]],
            figure4.panelEFGH[[4]][[1]],
            plot_label("Barrett's Esophagus from\nEsophageal Adenocarcinoma", 90, 6),
            figure4.panelIJKL[[1]][[1]],
            figure4.panelIJKL[[2]][[1]],
            figure4.panelIJKL[[3]][[1]],
            figure4.panelIJKL[[4]][[1]]
        ),
        guides = 'collect',
        tag_level = 'keep',
        design = "#ABCD\nEFGHI\nJKLMN\nOPQRS",
        heights = c(0.02, 1, 1, 1),
        widths = c(0.15, 1, 1, 1, 1)
    ) +
    patchwork::plot_annotation(
        tag_levels = list(list(
            "", "", "", "",
            "", "A", "B", "C", "D",
            "", "E", "F", "G", "H",
            "", "I", "J", "K", "L"
        ))
    )
ggplot2::ggsave(
    filename = "results/figures/figure4.pdf",
    plot = figure4,
    width = 17.4, height = 13, units = "cm",
    device = cairo_pdf
)
ggplot2::ggsave(
    filename = "results/figures/figure4.png",
    plot = figure4,
    width = 17.4, height = 13, units = "cm",
    dpi = 600
)
```

# Figure 5
## Figure 5 Panel A-D (CTLA, LAG4 expression)
```{r figure5.panelABD}
de <- de_genes_ea %>%
    dplyr::mutate(xstart = "CT", xend = "EA") %>%
    dplyr::bind_rows(
        de_genes_be %>%
            dplyr::mutate(xstart = "CT", xend = "BE")
    ) %>%
    dplyr::mutate(
        signif = dplyr::case_when(
            padj > 0.05 ~ "ns",
            padj > 0.01 ~ "*",
            padj > 0.001 ~ "**",
            TRUE ~ "***"
        )
    ) %>%
    dplyr::left_join(
        novel_dds %>%
            DESeq2::counts(normalized = TRUE) %>%
            apply(MARGIN = 1, FUN = max) %>%
            as.data.frame() %>%
            dplyr::rename(maxcnt = ".") %>%
            dplyr::mutate(gene_id = rownames(.)),
        by = "gene_id"
    )
genes <- c("S100A16", "FRAT2", "HHEX", "AIM2")
figure5.panelABCD <- lapply(genes, function(gene) {
    des <- de %>%
        dplyr::filter(gene_name == gene) %>%
        dplyr::mutate(
            ypos = max(maxcnt) * seq(from = 1.05, by = 0.1, length.out = nrow(.)),
            xstart = factor(xstart, levels = c("CT", "BE", "EA")),
            xend = factor(xend, levels = c("CT", "BE", "EA"))
        )
    plot <- novel_dds %>%
        DESeq2::counts(normalized = TRUE) %>%
        as.data.frame() %>%
        dplyr::mutate(gene_id = rownames(.)) %>%
        dplyr::left_join(
            novel_se %>%
                SummarizedExperiment::rowData() %>%
                as.data.frame() %>%
                dplyr::select(gene_id, gene_name, gene_biotype),
            by = "gene_id"
        ) %>%
        dplyr::filter(gene_name == gene) %>%
        tidyr::pivot_longer(
            dplyr::starts_with("SLDK"),
            names_to = "sample",
            values_to = "counts"
        ) %>%
        dplyr::left_join(
            novel_se %>%
                SummarizedExperiment::colData() %>%
                as.data.frame() %>%
                dplyr::select(sample, sample_id, condition_code),
            by = "sample"
        ) %>%
        ggplot2::ggplot(ggplot2::aes(x = condition_code, y = counts, fill = condition_code)) +
        ggplot2::geom_boxplot(outlier.size = 0.5) +
        ggplot2::geom_segment(
            data = des,
            ggplot2::aes(
                x = xstart, xend = xend, y = ypos, yend = ypos
            ),
            color = "black",
            inherit.aes = FALSE
        ) +
        ggplot2::geom_text(
            data = des,
            ggplot2::aes(
                x = (as.numeric(xend) - as.numeric(xstart))/2 + as.numeric(xstart),
                y = ypos,
                label = signif
            ),
            color = "black",
            size = 7,
            size.unit = "pt",
            hjust = 0.5,
            vjust = 0,
            inherit.aes = FALSE
        ) +
        ggplot2::scale_fill_manual(values = c(
            "CT" = "#666666",
            "BE" = "#2C9DEF",
            "EA" = "#D95F02"
        )) +
        theme_cell() +
        ggplot2::theme(legend.position = "none") +
        ggplot2::coord_cartesian(clip = "off") +
        ggplot2::labs(
            x = "Sample ID",
            y = "Counts",
            title = gene
        )
    panel = c(
        "S100A16" = "A",
        "FRAT2" = "B",
        "HHEX" = "C",
        "AIM2" = "D"
    )
    ggplot2::ggsave(
        paste0(
            "results/panels/figure5_panel",
            panel[[gene]],
            ".pdf"
        ),
        plot,
        units = "cm",
        height = 5,
        width = 4.35,
        device = cairo_pdf
    )
    return(list(plot))
}) %>%
do.call("c", args = .)
```


## Panel M: EA GSEA
```{r fig5.panel_e}
fgsea_ea <- fgsea::fgsea(
    pathways = fgsea::gmtPathways('data/references/hallmark_pathways.gmt'),
    stats = de_genes_ea %>%
        dplyr::filter(startsWith(gene_id, "ENSG")) %>%
        dplyr::summarize(
            .by = gene_name,
            stat = median(stat)
        ) %>%
        dplyr::arrange(dplyr::desc(stat)) %>%
        {
            dat = .
            res <- dat$stat %>%
                magrittr::set_names(dat$gene_name)
        }
)

figure5.panelE <- fgsea_ea %>%
    dplyr::filter(padj < 0.001) %>%
    dplyr::arrange(padj) %>%
    dplyr::slice_head(n = 20) %>%
    dplyr::mutate(
        title = stringr::str_replace(pathway, 'HALLMARK_', '') %>%
            stringr::str_replace_all('_', ' ')
    ) %>%
    dplyr::mutate(title = factor(title, levels = rev(title))) %>%
    ggplot2::ggplot(ggplot2::aes(x = NES, y = title, size = -log10(padj), color = NES)) +
    ggplot2::geom_segment(
        ggplot2::aes(y = title, yend = title, xend = NES, color = NES),
        x = 0, linewidth = 0.5, inherit.aes = FALSE, show.legend = FALSE
    ) +
    ggplot2::geom_point() +
    ggplot2::scale_x_continuous(
        limits = c(0, 1.05*max(fgsea_ea$NES)),
        expand = c(0,0)
    ) +
    ggplot2::scale_color_gradient2(
        breaks = c(1, 1.5, 2, 2.5, 3, 3.5),
        labels = c(1, 1.5, 2, 2.5, 3, 3.5),
        low = RColorBrewer::brewer.pal(n = 11, name = "PRGn")[1],
        mid = "white",
        high = RColorBrewer::brewer.pal(n = 11, name =  "PRGn")[11],
        name = "Normalized Enrichment Score"
    ) +
    ggplot2::scale_size_continuous(
        name = expression(-log[10](adjusted~p-value)),
        breaks = c(4, 8, 16, 32),
        labels = c(4, 8, 16, 32)
    ) +
    ggplot2::labs(
        x = "Normalized Enrichment Score",
        y = "Hallmark Pathway",
        title = "Esophageal Adenocarcinoma enriched pathways"
    ) +
    theme_cell() +
    ggplot2::theme(
        legend.position = 'right',
        axis.text.y = ggplot2::element_text(size = 5)
    ) +
    ggplot2::coord_cartesian(clip = 'off')
ggplot2::ggsave(
    filename = "results/panels/figure5_panelE.pdf",
    plot = figure5.panelE,
    width = 5, height = 4,
    device = cairo_pdf
)
ggplot2::ggsave(
    filename = "results/panels/figure5_panelE.png",
    plot = figure5_panelE,
    width = 5, height = 4, dpi = 600
)
```

## Panel F: BE GSEA
```{r fig4.panel_f}
fgsea_be <- fgsea::fgsea(
    pathways = fgsea::gmtPathways('data/references/hallmark_pathways.gmt'),
    stats = de_genes_be %>%
        dplyr::filter(startsWith(gene_id, "ENSG")) %>%
        dplyr::summarize(
            .by = gene_name,
            stat = median(stat)
        ) %>%
        dplyr::arrange(dplyr::desc(stat)) %>%
        {
            dat = .
            res <- dat$stat %>%
                magrittr::set_names(dat$gene_name)
        }
)

figure5.panelF <- fgsea_be %>%
    dplyr::filter(padj < 0.001) %>%
    dplyr::arrange(padj) %>%
    dplyr::slice_head(n = 20) %>%
    dplyr::mutate(
        title = stringr::str_replace(pathway, 'HALLMARK_', '') %>%
            stringr::str_replace_all('_', ' ')
    ) %>%
    dplyr::mutate(title = factor(title, levels = rev(title))) %>%
    ggplot2::ggplot(ggplot2::aes(x = NES, y = title, size = -log10(padj), color = NES)) +
    ggplot2::geom_segment(
        ggplot2::aes(y = title, yend = title, xend = NES, color = NES),
        x = 0, linewidth = 0.5, inherit.aes = FALSE, show.legend = FALSE
    ) +
    ggplot2::geom_point() +
    ggplot2::scale_x_continuous(
        limits = c(0, 1.05*max(fgsea_be$NES)),
        expand = c(0,0)
    ) +
    ggplot2::scale_color_gradient2(
        breaks = c(1, 1.5, 2, 2.5, 3, 3.5),
        labels = c(1, 1.5, 2, 2.5, 3, 3.5),
        low = RColorBrewer::brewer.pal(n = 11, name = "PRGn")[1],
        mid = "white",
        high = RColorBrewer::brewer.pal(n = 11, name =  "PRGn")[11],
        name = "Normalized Enrichment Score"
    ) +
    ggplot2::scale_size_continuous(
        name = expression(-log[10](adjusted~p-value)),
        breaks = c(4, 8, 16, 32),
        labels = c(4, 8, 16, 32)
    ) +
    ggplot2::labs(
        x = "Normalized Enrichment Score",
        y = "Hallmark Pathway",
        title = "Barrett's Esophagus enriched pathways"
    ) +
    theme_cell() +
    ggplot2::theme(
        legend.position = 'none',
        axis.text.y = ggplot2::element_text(size = 5)
    ) +
    ggplot2::coord_cartesian(clip = 'off')
ggplot2::ggsave(
    filename = "results/panels/figure5_panelF.pdf",
    plot = figure5.panelF,
    width = 5, height = 4,
    device = cairo_pdf
)
ggplot2::ggsave(
    filename = "results/panels/figure5_panelF.png",
    plot = figure5.panelF,
    width = 5, height = 4, dpi = 600
)
```

## Figure 5
```{r figure5}
figure5 <- patchwork::wrap_plots(c(
    figure5.panelABCD,
    patchwork::free(
        patchwork::wrap_plots(list(
            figure5.panelE,
            figure5.panelF
        ), ncol = 1, guides = "collect") +
        patchwork::plot_layout(heights = c(16, 12))
    )
)) +
    patchwork::plot_layout(design = "ABCD\nEEEE", heights = c(1, 3)) +
    patchwork::plot_annotation(tag_levels = 'A')
ggplot2::ggsave(
    "results/figures/figure5.pdf",
    figure5,
    units = "cm",
    width = 17.4,
    height = 15,
    device = cairo_pdf
)
ggplot2::ggsave(
    "results/figures/figure5.png",
    figure5,
    units = "cm",
    width = 17.4,
    height = 15,
    dpi = 600
)
```


# Supplementary figure 1

## Panel A: Quantification accuracy

```{r figS1.panel_a}
count_data <- combined_se %>%
    SummarizedExperiment::assay() %>%
    as.data.frame() %>%
    dplyr::mutate(gene_id = rownames(.)) %>%
    tidyr::pivot_longer(
        cols = dplyr::starts_with("SLDK"),
        names_to = "names",
        values_to = "counts"
    ) %>%
    dplyr::left_join(
        SummarizedExperiment::colData(combined_se) %>%
        as.data.frame(),
        by = c("names")
    ) %>%
    dplyr::select(-names) %>%
    tidyr::pivot_wider(
        names_from = transcriptome,
        values_from = counts
    ) %>%
    dplyr::left_join(
        biotypes,
        by = "gene_id"
    )
count_stats <- count_data %>%
    dplyr::filter(!startsWith(gene_id, "novel_gene_")) %>%
    dplyr::summarize(
        pearson_r = cor(`GENCODE`, `Novel cfRNA + GENCODE`, method = "pearson"),
        slope = coef(lm(`Novel cfRNA + GENCODE` ~ `GENCODE`))[2],
        intercept = coef(lm(`Novel cfRNA + GENCODE` ~ `GENCODE`))[1]
    )
density <- MASS::kde2d(
    x = count_data %>%
        dplyr::filter(!startsWith(gene_id, "novel_gene_")) %>%
        dplyr::pull(`GENCODE`) %>%
        magrittr::add(1) %>%
        log10(),
    y = count_data %>%
        dplyr::filter(!startsWith(gene_id, "novel_gene_")) %>%
        dplyr::pull(`Novel cfRNA + GENCODE`) %>%
        magrittr::add(1) %>%
        log10(),
    n = 75
)
figureS1.panelA <- data.frame(
        expand.grid(x=density$x, y=density$y),
        z = as.vector(density$z)
    ) %>%
    dplyr::mutate(z = z / max(.$z)) %>%
    ggplot2::ggplot(ggplot2::aes(x = x, y = y, z = z)) +
    ggplot2::geom_contour_filled(
        breaks = c(0, 0.0001, 0.001, 0.01, 0.1, 1)
    ) +
    ggplot2::geom_abline(
        slope = 1, intercept = 0, color = "white",
        alpha = 0.25, linetype = "dashed", linewidth = 0.5
    ) +
    ggplot2::annotate(
        "text", x = -Inf, y = Inf, size = 2, color = "white",
        hjust = -0.05, vjust = 1.2,
        label = paste0(
            "slope=", round(count_stats$slope, 3),
            "\nR^2=", round(count_stats$pearson_r**2, 3)
        )
    ) +
    ggplot2::labs(
        x = "GENCODE",
        y = "Novel cfRNA + GENCODE",
        fill = 'Density',
        title = "Quantification correlation"
    ) +
    ggplot2::scale_x_continuous(expand = c(0,0)) +
    ggplot2::scale_y_continuous(expand = c(0,0)) +
    viridis::scale_fill_viridis(
        option = 'D', discrete = TRUE,
        labels = c(
            expression("< " ~ 10^-4),
            expression(10^-4 ~ "-" ~ 10^-3),
            expression(10^-3 ~ "-" ~ 10^-2),
            expression(10^-2 ~ "-" ~ 10^-1),
            expression("> " ~ 10^-1)
        )
    ) +
    theme_cell() +
    ggplot2::theme(
        legend.background = ggplot2::element_blank(),
        legend.key.size = ggplot2::unit(0.25, "cm"),
    )
ggplot2::ggsave(
    filename = "results/panels/figureS1_panelA.pdf",
    plot = figureS1.panelA,
    width = 3, height = 3,
    device = cairo_pdf
)
ggplot2::ggsave(
    filename = "results/panels/figureS1_panelA.png",
    plot = figureS1.panelA,
    width = 3, height = 3, dpi = 600
)
```

## Panel B: Age+sex differences

```{r figS1.panel_b}
figureS1.panelB <- SummarizedExperiment::colData(novel_se) %>%
    as.data.frame() %>%
    dplyr::mutate(
        condition_code = factor(condition_code, levels = c('CT', 'BE', 'EA'))
    ) %>%
    ggplot2::ggplot(ggplot2::aes(x = condition_code, y = age, fill = condition_code)) +
    ggplot2::geom_boxplot(outlier.size = 0.5) +
    ggplot2::labs(
        x = "Sample Condition",
        y = "Patient age",
        title = "Sample ages"
    ) +
    theme_cell() +
    ggplot2::scale_fill_manual(values = c(
        'CT' = '#666666',
        'BE' = '#2C9DEF',
        'EA' = '#D95F02'
    )) +
    ggplot2::theme(legend.position = "none")
ggplot2::ggsave(
    'results/panels/figureS1_panelB.pdf',
    height = 3, width = 2, device = cairo_pdf
)
ggplot2::ggsave(
    'results/panels/figureS1_panelB.png',
    height = 3, width = 2, dpi = 600
)
```

## Panel C: ORF content

```{r figS1.panel_c}
figureS1.panelC <- readr::read_csv(
        'data/other/misc/orfs.csv',
        col_names = TRUE, show_col_types = FALSE,
        progress = FALSE
    ) %>%
    dplyr::select(aa_length) %>%
    dplyr::filter(aa_length > 50) %>%
    ggplot2::ggplot(ggplot2::aes(x = aa_length)) +
    ggplot2::geom_histogram(bins = 50, fill = "#2C9DEF") +
    ggplot2::labs(
        x = "Predicted ORF length (aa)",
        y = "Count",
        title = "Novel cfRNA encoded ORFs"
    ) +
    theme_cell() +
    ggplot2::lims(x=c(50,250)) +
    ggplot2::theme(legend.position = "none") +
    ggplot2::scale_y_continuous(expand = c(0,0))
ggplot2::ggsave(
    'results/panels/figureS1_panelC.pdf', figureS1.panelC,
    width = 3, height = 3, device = cairo_pdf
)
ggplot2::ggsave(
    'results/panels/figureS1_panelC.png', figureS1.panelC,
    width = 3, height = 3, dpi = 600
)
```

## Panel D: Conservation

```{r figS1.panel_d}
figureS1.panelD <- readr::read_tsv(
        'data/other/misc/phastcons.tsv', col_names = c('gene_id', 'score', 'txome'),
        show_col_types = FALSE, progress = FALSE
    ) %>%
    dplyr::arrange(score) %>%
    dplyr::mutate(
        txome = dplyr::case_when(
            txome == 'novel' ~ "Novel cfRNA",
            TRUE ~ "GENCODE"
        )
    ) %>%
    dplyr::mutate(
        .by = txome,
        ecdf = seq_along(score) / length(score)
    ) %>%
    ggplot2::ggplot(ggplot2::aes(x = score, y = ecdf, color = txome)) +
    ggplot2::geom_line(show.legend = FALSE) +
    ggplot2::geom_point(
        data = data.frame(
            txome = c("Novel cfRNA", "GENCODE"), score = c(0, 0), ecdf = c(0, 0)
        ), alpha = 0
    ) +
    theme_cell() +
    ggplot2::scale_color_manual(values = c(
        'Novel cfRNA' = '#2C9DEF',
        'GENCODE' = '#666666'
    ), guide = ggplot2::guide_legend(override.aes = list(shape = 15, size = 6, alpha = 1))) +
    ggplot2::labs(
        x = "PhastCons score",
        y = "ECDF",
        color = "Transcriptome",
        title = "Novel cfRNA conservation"
    ) +
    ggplot2::theme(legend.key.spacing.y = ggplot2::unit(-0.14, "cm"))
ggplot2::ggsave(
    'results/panels/figureS1_panelD.pdf', figureS1.panelD,
    width = 3, height = 3, device = cairo_pdf
)
ggplot2::ggsave(
    'results/panels/figureS1_panelD.png', figureS1.panelD,
    width = 3, height = 3, dpi = 600
)
```

## Panel E: Genome Browser

```{r figS1.panelE}
figureS1.panelE <- magick::image_read_pdf('results/panels/figureS1_panelE.pdf', pages=1) %>%
        magick::image_crop(geometry='2700x1350', gravity = 'NorthWest') %>%
        grid::rasterGrob() %>%
        ggplotify::as.ggplot() +
        ggplot2::theme_void() +
        ggplot2::labs(title = "Supporting reads for an example novel transcript") +
        ggplot2::theme(
            plot.title = ggplot2::element_text(
                size = 6, color = "#282A36", face = 'bold',
                hjust = 0.5, vjust = 1.2
            ),
            plot.tag = ggplot2::element_text(
                size = 10, color = "#282A36", face = 'bold'
            )
        )
```

## Figure S1

```{r figS1}
figureS1 <- patchwork::wrap_plots(list(
            figureS1.panelA,
            figureS1.panelB,
            figureS1.panelC,
            figureS1.panelD,
            patchwork::free(figureS1.panelE)
        ),
        design = "ABCD\nEEEE",
        widths = c(1.25,1,1,1.25),
        heights = c(1.75,5),
        guides = 'keep'
    ) +
    patchwork::plot_annotation(tag_levels = 'A')
ggplot2::ggsave(
    'results/figures/figureS1.pdf', figureS1,
    width = 17.4, height = 13.5, units = "cm",
    device = cairo_pdf
)
ggplot2::ggsave(
    'results/figures/figureS1.png', figureS1,
    width = 17.4, height = 13.5, units = "cm",
    dpi = 600
)
```
